MSB1015 Project

### 2024-10-14 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Added more comments and scaled for the initial PCA

# 0. Setup
```{r setup, include=FALSE}
library(rstudioapi)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(GGally)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()

data <- read_csv("METABRIC_RNA_Mutation.csv")
```

@@

# Check for missing values
missing_counts <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
  filter(missing_count > 10)

# Plot where those missing values were found
ggplot(missing_counts, aes(x = reorder(column, missing_count), y = missing_count)) +
  geom_bar(stat = "identity", fill = "skyblue1") +
  coord_flip() +
  labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values")

data$tumor_stage <- as.factor(data$tumor_stage)

data$neoplasm_histologic_grade <- as.factor(data$neoplasm_histologic_grade)

# Apply the modifications
data_filled <- data %>%
  # 1. Replace missing values with "Unknown" for all categorical columns
  mutate(
    oncotree_code = replace_na(oncotree_code, "Unknown"),
    cancer_type_detailed = replace_na(cancer_type_detailed, "Unknown")
  ) %>%
  # 3. Replace missing values with the median for "tumor_size"
  mutate(tumor_size = ifelse(is.na(tumor_size), median(tumor_size, na.rm = TRUE), tumor_size)) %>%
  # 4. Replace missing values with 0 for columns ending with "_mut"
  mutate(across(ends_with("_mut"), ~ replace_na(., 0))) %>%
  # 5. Convert the columns ending with "_mut" to a logical vector
  mutate(across(ends_with("_mut"), ~ ifelse(. != "0", 1, 0))) %>%
  # 6.
  rowwise() %>%
  mutate(mutation_count = sum(across(ends_with("_mut")) == 1))

@@

# Perform PCA to check the columns with the highest amount of missing values
pca <- prcomp(pca_gx)

@@

# 0.2 Imputation (this step will take some time)
```{r impute, echo=FALSE}
# Load necessary libraries
library(mice)

# Get rid of unwanted characters from the column names
colnames(data_filled) <- make.names(colnames(data_filled), unique = TRUE)

@@

# Identify categorical variables
fac_vars <- names(data_filled)[sapply(data_filled, is.factor)]

# Specify methods for all variables, setting CART for categorical variables
meth <- make.method(data_filled)
meth[fac_vars] <- "cart" # Use CART for all categorical variables

@@

imputed_data <- complete(imputation_result)

write.csv(imputed_data, file = "imputed_data.csv", row.names = FALSE)
```

@@

# 2. The random forest
```{r data tranformation for random forest, echo=FALSE}
# Load required packages
library(randomForest)
library(caTools)
library(caret)
library(nestedcv)
library(mlbench)
library(pbapply)
library(ROCR)
library(PRROC)
library(pROC)
library(gridExtra)

# Create a function to form the prognosis variable
categorise_prognosis <- function(months) {
  case_when(
    months <= 60 + (60 * 0.25) ~ "will_die_within_5_years",
    months > 60 + (60 * 0.25) ~ "will_die_within_10_years",
  )
}

# Create the categorical dataframe
categorical_data <- data_filtered %>%
  select(age_at_diagnosis:death_from_cancer) %>% # Extract the categorical variables
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = as.factor(categorise_prognosis(overall_survival_months)) # Call the function to create 'prognosis'
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
  as.data.frame()

source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed

@@

# 2.1 The random forest function
```{r random forest function, echo=FALSE}
random_forest <- function(data, ntree, nrun) {
  # Set the seed for reproducibility using the L'Ecuyer-CMRG method
  set.seed(111, "L'Ecuyer-CMRG")

@@

  # Compute the average across the rows
  t_final_metrics_df <- t_final_metrics_df %>%
    rowwise() %>%
    mutate(average = mean(c_across(everything()), na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(Metric = c("AUC", "Balanced Accuracy"))

@@

# 4. ROC and PR curves on the best model from the best class
```{r accuracy and plotting, echo=FALSE}
# Calculate accuracy on the train set for the best model
pred_table_training <- table(best_class_data$best_model_overall$output$predy, best_class_data$best_model_overall$output$testy)
train_confusion <- confusionMatrix(pred_table_training)
train_confusion

# Call the plotting function for the confusion matrix
source("~/University/MSB/MSB1015/confusionPlot.R") # Change the directory as needed
confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)")

# Get predicted probabilities for the positive class ("will_die_within_10_years")
pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$train, type = "prob")
positive_class_probs <- pred_probs[, "will_die_within_10_years"]

# Get the precision recall values
best_class_data$best_model_overall$prc <- prc(best_class_data$best_model_overall)

@@

# Plot precision and recall
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC Training", las = 1)
legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 3)), bty = "n")
plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-recall Training")
legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 3)), bty = "n")

@@

# Extract the importance of the variables for the best model
importance <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$importance) %>%
  mutate(
    will_die_within_10_years = will_die_within_10_years * 10,
    will_die_within_5_years = will_die_within_5_years * 10,
    column = row.names(best_class_data$best_model_overall$final_fit$finalModel$importance)
  )

importance_sorted <- importance[order(importance$will_die_within_10_years, decreasing = TRUE), ]
head(importance_sorted)

# Plot the importance of the variables
ggplot(importance, aes(x = reorder(column, will_die_within_10_years))) +
  geom_bar(aes(y = will_die_within_10_years, fill = "Will die within 10 years"),
    position = "dodge", stat = "identity"
  ) +
  geom_bar(aes(y = will_die_within_5_years, fill = "Will die within 5 years"),
    position = "dodge", stat = "identity"
  ) +
  scale_fill_manual(values = c(
    "Will die within 10 years" = "skyblue1",
    "Will die within 5 years" = "tomato"
  )) +
  coord_flip() +
  labs(x = "Feature", y = "Importance Score", title = "Variable Importance", fill = "Prognosis") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Extract the error rates (OOB)
err_rate <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$err.rate) %>%
  mutate(ntrees = 1:nrow(best_class_data$best_model_overall$final_fit$finalModel$err.rate))

# Plot OOB
ggplot(err_rate, aes(x = ntrees, y = OOB)) +
  geom_line(color = "skyblue1", linewidth = 1) +
  labs(x = "Number of Trees", y = "OBB Error", title = "Out Of Bags (OOB) error") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

# 5. Testing on the best model
```{r testing, echo=FALSE}
# Get predicted probabilities for the positive class ("will_die_within_10_years")
pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$test, type = "prob")
positive_class_probs <- pred_probs[, "will_die_within_10_years"]

# Generate the ROC curve
roc_curve <- roc(best_class_data$test$prognosis, positive_class_probs, levels = levels(best_class_data$test$prognosis), direction = ">")

# Create Precision-Recall curve
pr_curve <- pr.curve(
  scores.class0 = positive_class_probs,
  weights.class0 = as.numeric(best_class_data$test$prognosis == "will_die_within_10_years"),
  curve = TRUE
)

# Save precision and recall plots as PNG
png("precision_recall_plots_train_test.png", width = 1600, height = 800)

# Plot combined graphs
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC", las = 1)
plot(roc_curve, col = "blue", add = TRUE, main = "ROC", las = 1)
legend("bottomright",
  legend = c(
    paste0("AUC (Training) = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 2)),
    paste0("AUC (Testing) = ", signif(roc_curve$auc, 2))
  ),
  col = c("red", "blue"), lty = 1, bty = "n"
)

# Plot the Precision-Recall curve for the training data
plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision")
lines(pr_curve$curve[, 3], pr_curve$curve[, 1], col = "blue")
legend("topright",
  legend = c(
    paste0("AUC (Training) = ", round(best_class_data$best_model_overall$prc$auc, 2)),
    paste0("AUC (Testing) = ", round(pr_curve$auc.integral, 2))
  ),
  col = c("red", "blue"),
  lty = 1,
  bty = "n"
)

# Reset plotting parameters and close PNG
par(op)
dev.off()

# Find the best threshold from the ROC curve (maximizing sensitivity-specificity balance)
best_threshold <- as.numeric(coords(roc_curve, "best", ret = "threshold"))
print(paste("Best threshold based on ROC curve:", best_threshold))

# Apply the best threshold for classification
pred_best_threshold <- ifelse(positive_class_probs > best_threshold,
  "will_die_within_10_years",
  "will_die_within_5_years"
)

# Ensure the predicted classes are factors with the correct levels
pred_best_threshold <- factor(pred_best_threshold, levels = levels(best_class_data$test$prognosis))

# Calculate accuracy on the test set with the best threshold
pred_table <- table(pred_best_threshold, best_class_data$test$prognosis)
dimnames(pred_table) <- list(
  c("will_die_within_10_years", "will_die_within_5_years"),
  c("will_die_within_10_years", "will_die_within_5_years")
)
test_confusion <- confusionMatrix(pred_table)
test_confusion

confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")
```

>>>>

# 0. Setup
```{r setup, include=FALSE}
# List of required packages
required_packages <- c(
  "rstudioapi", "readr", "dplyr", "ggplot2", "tidyr",
  "stringr", "GGally", "mice", "isotree", "randomForest",
  "caTools", "caret", "nestedcv", "mlbench", "pbapply",
  "ROCR", "PRROC", "pROC", "gridExtra"
)

# Check and install missing packages
new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(new_packages)) {
  install.packages(new_packages)
}

# Load libraries
library(rstudioapi)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(GGally)
library(mice)
library(randomForest)
library(caTools)
library(caret)
library(nestedcv)
library(mlbench)
library(pbapply)
library(ROCR)
library(PRROC)
library(pROC)
library(gridExtra)

# Set the working directory to the one where this code is
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()

# Load the data
data <- read_csv("METABRIC_RNA_Mutation.csv")
```

@@

# Check for missing values
missing_counts <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>% # Summarise missing values per column
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>% # Reshape into long format
  filter(missing_count > 10) # Filter columns with more than 10 missing values

# Plot where those missing values were found
ggplot(missing_counts, aes(x = reorder(column, missing_count), y = missing_count)) +
  geom_bar(stat = "identity", fill = "royalblue") +
  coord_flip() +
  labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values")

# Convert specific columns to factors
data$tumor_stage <- as.factor(data$tumor_stage)
data$neoplasm_histologic_grade <- as.factor(data$neoplasm_histologic_grade)

# Apply the modifications
data_filled <- data %>%
  # 1. Replace missing values with "Unknown" for specific categorical columns (not suitable for imputation)
  mutate(
    oncotree_code = replace_na(oncotree_code, "Unknown"),
    cancer_type_detailed = replace_na(cancer_type_detailed, "Unknown")
  ) %>%
  # 2. Replace missing values in "tumor_size" with the median value of the column
  mutate(tumor_size = ifelse(is.na(tumor_size), median(tumor_size, na.rm = TRUE), tumor_size)) %>%
  # 3. Replace missing values with 0 for all columns that end with "_mut"
  mutate(across(ends_with("_mut"), ~ replace_na(., 0))) %>%
  # 4. Convert values in columns ending with "_mut" to a binary vector (values not equal to "0" are converted to 1)
  mutate(across(ends_with("_mut"), ~ ifelse(. != "0", 1, 0))) %>%
  # 5. Calculate the "mutation_count" by summing the binary values (1s) in columns that end with "_mut" for each row
  rowwise() %>%
  mutate(mutation_count = sum(across(ends_with("_mut")) == 1))

@@

# Perform PCA to check the columns with the highest amount of missing values
pca <- prcomp(pca_gx, center = TRUE, scale. = TRUE)

@@

# 0.2 Imputation (this step will take some time)
```{r impute, echo=FALSE}
# Get rid of unwanted characters from the column names (the sign '+' from one of the variables)
colnames(data_filled) <- make.names(colnames(data_filled), unique = TRUE)

@@

@@

# Collect the results
imputed_data <- complete(imputation_result)

# Save the imputed data in the working directory
write.csv(imputed_data, file = "imputed_data.csv", row.names = FALSE)
```

@@

# 2. The random forest
```{r data tranformation for random forest, echo=FALSE}
# Create a function to form the prognosis variable based on overall survival months
categorise_prognosis <- function(months) {
  case_when(
    months <= 60 + (60 * 0.25) ~ "will_die_within_5_years",
    months > 60 + (60 * 0.25) ~ "will_die_within_10_years"
  )
}

# Create the categorical dataframe
categorical_data <- data_filtered %>%
  select(age_at_diagnosis:death_from_cancer) %>% # Extract the categorical variables
  filter(death_from_cancer != "Died of Other Causes") %>% # Remove records for non-cancer deaths
  filter(overall_survival == 0) %>% # Keep only the rows where overall survival is 0
  mutate(
    prognosis = as.factor(categorise_prognosis(overall_survival_months)) # Apply function to create 'prognosis'
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>% # Drop unnecessary columns
  as.data.frame() # Convert to a dataframe

# Load the function for PCA
source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed

@@

# 2.1 The random forest function
```{r random forest function, echo=FALSE}
# Create the function to perform random forest classification
random_forest <- function(data, ntree, nrun) {
  # Set the seed for reproducibility using the L'Ecuyer-CMRG method
  set.seed(111, "L'Ecuyer-CMRG")

@@

  # Compute the average across the rows
  t_final_metrics_df <- t_final_metrics_df %>%
    rowwise() %>% # Perform row-wise operations
    mutate(average = mean(c_across(everything()), na.rm = TRUE)) %>% # Calculate row-wise average
    ungroup() %>%
    mutate(Metric = c("AUC", "Balanced Accuracy")) # Add a Metric column

@@

# 4. ROC and PR curves on the best model from the best class
```{r accuracy and plotting, echo=FALSE}
# Calculate accuracy on the train set for the best model
pred_table_training <- table(best_class_data$best_model_overall$output$predy, best_class_data$best_model_overall$output$testy) # Create prediction table
train_confusion <- confusionMatrix(pred_table_training) # Compute confusion matrix
train_confusion

# Call the plotting function for the confusion matrix
source("~/University/MSB/MSB1015/confusionPlot.R") # Load external plot function (update directory if needed)
confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)") # Plot confusion matrix

# Get predicted probabilities for the positive class ("will_die_within_10_years")
pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$train, type = "prob") # Predict probabilities
positive_class_probs <- pred_probs[, "will_die_within_10_years"] # Extract probabilities for "will_die_within_10_years"

# Get the precision-recall values
best_class_data$best_model_overall$prc <- prc(best_class_data$best_model_overall) # Calculate precision-recall

@@

# Plot precision and recall
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1) # Set up the plotting layout
plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC Training", las = 1) # Plot ROC curve
legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 3)), bty = "n") # Add AUC legend
plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-recall Training") # Plot Precision-Recall curve
legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 3)), bty = "n") # Add AUC legend for PRC

@@

# Extract the importance of the variables for the best model
importance <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$importance) %>% # Get variable importance
  mutate(
    will_die_within_10_years = will_die_within_10_years * 10, # Scale the importance scores
    will_die_within_5_years = will_die_within_5_years * 10,
    column = row.names(best_class_data$best_model_overall$final_fit$finalModel$importance) # Extract feature names
  )

importance_sorted <- importance[order(importance$will_die_within_10_years, decreasing = TRUE), ] # Sort by importance
head(importance_sorted)

# Plot the importance of the variables
ggplot(importance, aes(x = reorder(column, will_die_within_10_years))) +
  geom_bar(aes(y = will_die_within_10_years, fill = "Will die within 10 years"),
    position = "dodge", stat = "identity"
  ) +
  geom_bar(aes(y = will_die_within_5_years, fill = "Will die within 5 years"),
    position = "dodge", stat = "identity"
  ) +
  scale_fill_manual(values = c(
    "Will die within 10 years" = "skyblue1",
    "Will die within 5 years" = "tomato"
  )) +
  coord_flip() +
  labs(x = "Feature", y = "Importance Score", title = "Variable Importance", fill = "Prognosis") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) # Create variable importance plot

# Extract the error rates (OOB)
err_rate <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$err.rate) %>% # Extract error rates
  mutate(ntrees = 1:nrow(best_class_data$best_model_overall$final_fit$finalModel$err.rate)) # Add number of trees

# Plot OOB error
ggplot(err_rate, aes(x = ntrees, y = OOB)) +
  geom_line(color = "skyblue1", linewidth = 1) +
  labs(x = "Number of Trees", y = "OOB Error", title = "Out Of Bag (OOB) Error") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) # Plot OOB error
```

# 5. Testing on the best model
```{r testing, echo=FALSE}
# Get predicted probabilities for the positive class ("will_die_within_10_years")
pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$test, type = "prob")
positive_class_probs <- pred_probs[, "will_die_within_10_years"] # Select probabilities for "will_die_within_10_years"

# Generate the ROC curve for the test set
roc_curve <- roc(best_class_data$test$prognosis, positive_class_probs, levels = levels(best_class_data$test$prognosis), direction = ">")

# Create Precision-Recall curve for the test set
pr_curve <- pr.curve(
  scores.class0 = positive_class_probs, # Probabilities for the positive class
  weights.class0 = as.numeric(best_class_data$test$prognosis == "will_die_within_10_years"), # Binary vector for positive class
  curve = TRUE # Return curve for plotting
)

# Save precision and recall plots as PNG
png("precision_recall_plots_train_test.png", width = 1600, height = 800) # Save output to PNG file

# Plot combined graphs (ROC and Precision-Recall curves side by side)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1) # Set layout to 1 row, 2 columns

# Plot ROC curve for both training and testing
plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC", las = 1) # Training ROC
plot(roc_curve, col = "blue", add = TRUE, main = "ROC", las = 1) # Testing ROC
legend("bottomright", # Add legend with AUC values
  legend = c(
    paste0("AUC (Training) = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 2)),
    paste0("AUC (Testing) = ", signif(roc_curve$auc, 2))
  ),
  col = c("red", "blue"), lty = 1, bty = "n"
)

# Plot Precision-Recall curve for both training and testing
plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision") # Training PRC
lines(pr_curve$curve[, 3], pr_curve$curve[, 1], col = "blue") # Testing PRC
legend("topright", # Add legend with AUC values
  legend = c(
    paste0("AUC (Training) = ", round(best_class_data$best_model_overall$prc$auc, 2)),
    paste0("AUC (Testing) = ", round(pr_curve$auc.integral, 2))
  ),
  col = c("red", "blue"),
  lty = 1,
  bty = "n"
)

# Reset plotting parameters and close PNG file
par(op)
dev.off()

# Find the best threshold from the ROC curve
best_threshold <- as.numeric(coords(roc_curve, "best", ret = "threshold")) # Threshold maximizing sensitivity-specificity
print(paste("Best threshold based on ROC curve:", best_threshold))

# Apply the best threshold for classification
pred_best_threshold <- ifelse(positive_class_probs > best_threshold, "will_die_within_10_years", "will_die_within_5_years") # Classify based on threshold

# Ensure predicted classes have the correct factor levels
pred_best_threshold <- factor(pred_best_threshold, levels = levels(best_class_data$test$prognosis))

# Calculate accuracy on the test set using the best threshold
pred_table <- table(pred_best_threshold, best_class_data$test$prognosis) # Create confusion matrix table
dimnames(pred_table) <- list( # Assign dimension names to match factor levels
  c("will_die_within_10_years", "will_die_within_5_years"),
  c("will_die_within_10_years", "will_die_within_5_years")
)
test_confusion <- confusionMatrix(pred_table) # Compute confusion matrix
test_confusion

# Plot the confusion matrix for the test set
confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")
```
